---
title: "Lesson 0 - Introduction to R"
---


R is a high-level programming language commonly used for data analysis and data visualization.

Additionally, R is also an interpreted language. This means that the R interpreter reads a line of the program being executed, checks for incorrect syntax that will produce an error, translates R syntax into machine code (0's & 1's) to be executed by the computer, and prints the output of that line before proceeding to the next line.


```{r}
#In R, comments are created by adding the hash tag before your sentence so that the interpreter  knows that they are comments.

print("Hello World!") #Interpreter reads and translates this line first
print("My name is _") #Then the interpreter reads and translates this line next
```


Because each line is translated and executed one line at a time, if the interpreter detects a syntax error, it will stop executing the program and produce an error. Any code under the line producing the error will not be executed.


```{r}
print("Hello World!") 
# This line will produce an error. Note how the first line is printed but neither the second nor third line are.
print("Incorrect syntax" 
print("My name is _") 
```


