---
title: "Lesson 4 - Basic Data Analysis & Data Visualization"
---


```{r}
data <- read.csv("C:/Users/donis/Documents/R Workshop/iris.csv")
```


`R` has base functions to conduct regression and ANOVA analyses.

For linear regression you can use `R`'s built-in `lm()` function. The form for the regression equation is `Dependent Variable ~ Independent Variable`. Also, you need to use the `data =` argument so that `lm()` knows that you are referring to variable names inside your data frame. You can see the results of the regression model using `R`'s built-in `summary()`function. 

`Pr(>|t|)` is the p-value if you see an asterisk(`*`) next to the number, your variable is significant.

```{r}
#`lm()` outputs an `lm` object that you can save inside a variable
mod_1 <-  lm(Sepal.Length ~ Petal.Width, data = data)
# You can see
summary(model_1)
```

Here, we add an additional variable to the model.
```{r}
mod_2 <-  lm(Sepal.Length ~ Petal.Width + Petal.Length, data = data)

summary(model_2)
```

We can use `R`'s built-in `anova()` function to see which model is better (reduces the residuals).
```{r}
anova(model_1,model_2)
```

```{r}
plot(model_1)
```



```{r}
hist(resid(model_1),main = "Histogram", xlab = "Residuals")
```



```{r}
aov_model <- aov(Sepal.Length ~ Species, data = data)

summary(aov_model)
```

```{r}
model_3 <- lm(Sepal.Length ~ Species, data = data)

summary(model_3)
```
To prove that the intercept is the mean sepal length of the setosa species, we will use `which()` to get the indices of data$Species that is equal to setosa. The `which()` function is useful for obtaining the indices of values in a vector that meet a logical condition. The individual columns in a data frame can be treated as a vector since it is a one dimensional object.


```{r}
which(data$Species == "setosa")
```
We can use these indices to get the rows of the "Sepal Length" column for only the "setosa" species. Then obtain the mean.

```{r}
mean(data[which(data$Species == "setosa"),"Sepal.Length"])
```
We can also get the difference in the means of the other species and the "setosa" species.
```{r}
mean(data[which(data$Species == "versicolor"),"Sepal.Length"]) - mean(data[which(data$Species == "setosa"),"Sepal.Length"]) 
```

```{r}
mean(data[which(data$Species == "virginica"),"Sepal.Length"]) - mean(data[which(data$Species == "setosa"),"Sepal.Length"]) 
```


We can use `R`'s built-in `relevel()` function to change the reference group for our regression model.
```{r}
levels(data$Species)

data$Species <- relevel(data$Species, ref = "virginica")

levels(data$Species)

```






```{r}
#reduced_data <-
data %>% group_by(Species) %>% 
  filter(Species %in% c("versicolor", "virginica")) %>% mutate(Species = relevel(c("versicolor" = 0, "virginica" = 1)))

#binary_model <- glm(Species ~ Sepal.Length, family = "binomial")



```






```{r}
model_3 <- lm(Sepal.Length ~ Species, data = data)

summary(model_3)
```




# Plot a scatter plot
plot(data$independent_variable, data$dependent_variable,
     xlab = "Independent Variable", ylab = "Dependent Variable", 
     main = "Scatter Plot of Independent vs. Dependent Variable")

# Plot a scatter plot using ggplot2
ggplot(data, aes(x = independent_variable, y = dependent_variable)) +
  geom_point() +
  labs(x = "Independent Variable", y = "Dependent Variable", 
       title = "Scatter Plot of Independent vs. Dependent Variable")