---
title: "Lesson 2 - Importing & Exporting Data in R"
---

In this lesson, we will cover how to import and export data with common file extensions `.csv`,`.tsv`, & `.xlsx` into `R`.

## Importing Data

`R` has built-in functions to import export comma-separated values(csv) and tab-separated values(tsv) files. 

To import csv files into `R`, you can use `read.csv()` and to import tsv files, you can use `read.table()`.

Both functions have multiple arguments; however, some arguments that are most commonly used are:

- `file =`: To specify the location of the file that you are importing. You can use the absolute path location (full path from the root) to your file or the relative path location (location relative to your current working directory).

- `header =`: This argument only accepts logical arguments (TRUE or FALSE) and defaults to `header = TRUE`. If set to TRUE, the first line in your will be used as the column names of your data frame. If set to FALSE, `R` will use default column names for your data frame.

- `sep =`: This argument specifies the deliminator, a character that separates the values in your file. For `read.csv()` this argument defaults to `sep = ","` and for read.table() this argument defaults to `sep = ""` (which will recognize values separated by one or two spaces and tabs). You usually won't need to change this parameter, unless the deliminator is not separated by the character in the default argument.

Note: Each argument must be separated by a comma (`,`).

You can check your current working directory by using `getwd()` if you want to use a relative path to import your data. 

```{r}
getwd()
```
You can change your working directory using, `setwd()`

```{r}
setwd("C:/Users/Documents/")
getwd()
```

#### Importing a csv file:
```{r}
#Importing from absolute path location
data <- read.csv(file = "C:/Users/Documents/iris.csv", header = T, sep = ",")

#If your file is located in your current working directory, you can simply use the name of your file. This would be the relative path location of your file.
data <- read.csv(file = "iris.csv", header = T, sep = ",")
```


#### Importing a tsv file:
```{r}
data <- read.table(file = "C:/Users/Documents/iris.tsv", header = T, sep = "\t")
```

#### Importing an Excel file:

R does not have a built-in function to read Excel files. 

In R, you can download and install packages that will allow you to use additional useful functions, using the install.packages().The `install.packages()` function has a `repos =` argument to specify the download location of the package you wish to install. This argument automatically defaults to downloading packages from Comprehensive R Archive Network (CRAN) (`repos = "http://cran.us.r-project.org"`), which is the main repository for R packages. The majority of the R packages that you will need will already be on CRAN. You can go [here](http://lib.stat.cmu.edu/R/CRAN/) if you wish to see all of the packages on CRAN with descriptions of what those packages are for. Simply use `Ctrl + F` to find paclkages with certain keywords in their descriptions.

A good package to import Excel files into `R` is `readxl`.

`readxl` uses the `path =` as the argument to specify your file's location.

If you want to check the complete list of functions,data sets,and other content available inside a package, use `library(help = "package_name")` (ex. `library(help = "readxl")`). You will be presented with a page documenting all the contents within a package.



```{r}
#Name of the package must be in quotes
install.packages("readxl") 
```



To use the functions in a package, you can either:

- Load it into `R`'s search path (a list of directories where `R` searches for packages and files) using `library()`. The `library()` function adds the package directory into the search path so the contents within the package can be accessed.


```{r}
#Name of the package should not be in quotes
library(readxl)
# The argument to specify your file location `path = `
data <- read_xlsx(path = "C:/Users/Documents/iris.xlsx")

#Using `R`'s `search()` function shows that the package is now in the search path with the other base `R` packages
#that are automatically loaded in when `R` is activated
search()
```



- Access the contents in a package using the double colon operator(`::`)
```{r}
data <- readxl::read_xlsx(path = "C:/Users/Documents/iris.xlsx")
```



## Exporting Data

R has built-in functions to save csv and tsv files.

To save csv files, you can use `write.csv()`, and the save tsv files you can use `write.tsv()`.

The most common arguments that you will use in these functions are:

- `x = ` : Specify the name of the object (variable) containing the data that you wish to save.

- `file = ` : To specify the location that you want your file to be saved and to specify the name of your file.

- `sep = ` : Specify what you want the values in you file to be separated by. For `write.csv()`, this argument defaults to `sep = ","` and will ignore any changes to this argument that you attempt make. 

#### Exporting csv file:
```{r}
write.csv(x = data, file = "C:/Users/Documents/iris.csv")
```

#### Exporting tsv file:


```{r}
write.table(x = data, file = "C:/Users/Documents/iris.tsv", sep = "\t")
```


#### Exporting Excel file:

R does not have a built-in function to export Excel files. A good package to use is `xlsx`. This package requires [Java](https://www.java.com/en/download/) to be installed on your operating system to work.

`xlsx` also uses the `x =` and `file =` arguments but does not have a `sep =` argument.


```{r}
install.packages("xlsx")
library(xlsx)

write.xlsx(x = data, file = "C:/Users/Documents/iris.xlsx")
```


*Note: Sometimes `R` packages may conflict, you can unload a package from `R`'s search path using `detach()`.*

```{r}
# Detaching packages requires that you add "package:" before the name of the package
detach("package:readxl")
#It is now detached from the search path
search()
```


